A tool set for random number generation on GPUs in R

### 2021-11-10 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.Rnw
paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@2955 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

An important way of parallel computing with \proglang{R} is using Graphics Processing Units (GPUs). GPUs can perform thousands of tests simultaneously, which makes them powerful at doing massive parallel computing and they are relatively cheap compared to multicore CPU's. Although there has been a few \proglang{R} packages developed that provide some GPU capability in the past years, they come with some restrictions. Packages such as \pkg{gputools}, \pkg{gpumatrix}, \pkg{cudaBayesreg}, \pkg{rpud} (available on github), are now no longer maintained, the popular \pkg{tensorflow} \citep{tensorflow1} package uses GPU via \proglang{Python}, which makes it hard to be incorporated in a new \proglang{R} package that uses GPU. All of these mentioned packages are restricted to \proglang{R} users with NVIDIA GPUs. \pkg{gpuR} \citep{gpur1} is the only \proglang{R} package with a convenient interface between \proglang{R} and GPUs and it is created for any \proglang{R} user with a GPU device. By utilizing the \pkg{ViennaCL} \citep*{rupp2016viennacl} library, it provides a bridge between \proglang{R} and non-proprietary GPUs through the \proglang{OpenCL} (Open Computing Language) backend, which when combined with \pkg{Rcpp} \citep{rcpp1} gives a building block for other \proglang{R} packages.

Random number generation is critical in simulation-based statistical inference, machine learning and many other scientific fields. While most random number generators are sequential, the \proglang{R} packages \pkg{parallel} (not available on cran), \pkg{rlecuyer} \citep{sevcikova2015package} and \pkg{future} \citep*{future1.19.1} are able to generate random numbers in parallel on multicore CPUs. More specifically, \pkg{parallel} writes an interface for the \pkg{RngStreams}, a \proglang{\texttt{C++}} library by \cite{l2002object} which is based on a combined multiple-recursive generator (MRG). \pkg{future} also uses the combined MRG algorithm for generating random numbers, however, it is not very clear how the streams are implemented. For up-to-date review papers on the generation of random numbers on parallel devices, and GPUs in particular, see:  \cite{rLEC15a,rLEC17p,rLEC21a}.

There is not yet an \proglang{R} package that is able to generate random numbers on GPU, this is complicated to do because multiple streams will be needed for parallel generation of random numbers, and the stream states need to be able to be restored for reproducibility of results, which is essential in many simulation applications.  We introduce the \proglang{R} package \pkg{clrng} that leverages the \pkg{gpuR} package, and can generate random numbers on GPU by using the \pkg{clRNG} \citep{l2015clrng} library, and can thus accelerate several types of statistical simulation and modelling.

@@

\section{Uniform random number generation} \label{}
Uniform random number generators (RNGs) are essential in simulating random numbers from all types of probability distributions. \cite{l2012random} summarized the usual two steps to generate a random variable in computational statistics: (1) generating independent and identically distributed (i.i.d.) uniform random variables on the interval $(0, 1)$, (2) applying transformations to these i.i.d. $U(0, 1)$ random variables to get samples from the desired distribution. \cite{l2012random} and \cite{robert2004random} present many general transformation methods for generating non-uniform random variables, for example, the most frequently used inverse transform method, the Box-Muller algorithm \citep{box1958note} for Gaussian random variable generation, and so on, which are all built on uniform random variables. %In the rest paper, ``RNG'' refers to uniform pseudo random number generator.

\pkg{clRNG} is an \proglang{OpenCL} library for uniform random number generation, it provides four different RNGs: the MRG31k3p, MRG32k3a, LFSR113, and Philox-4×32-10 generators. These four RNGs use different types of constructions. \pkg{clrng} package uses the MRG31k3p RNG, making it able to generate random numbers on GPUs. We choose MRG31k3p as the base generator for the following reasons:  the original \pkg{RNGStreams} package \citep{l2002object} was built with MRG32k3a, which was designed to be implemented in ``double'' in floating point(?), and not with 32-bit integers. The MRG31k3p generator was designed later, specifically for 32-bit integer arithmetic, so it runs faster on the 32-bit GPUs.  It is also faster than Philox-4×32-10 \citep{rLEC00b}. MRG31k3p was also statistically tested extensively and successfully, \cite[see][]{rLEC07b}.

@@

\subsection{Creating streams}\label{createstreams}
The popular way of obtaining multiple streams is to take an RNG with a long period and cut the RNG's sequence into very long pieces, these pieces are adjacent and of equal length $Z$, so each piece can be used as a separate stream. Creating a new stream amounts to computing its starting point (seed). Each of these streams can also be partitioned into substreams with equally-spaced starting points \citep{l2002object, rLEC15a}, which are occasionally useful.

@@

In \proglang{clRNG} library, for the MRG31k3p RNG, the entire period length of approximately $2^{185}$ is divided into approximately $2^{51}$ non-overlapping streams of length $Z = 2^{134}$. Each stream is further partitioned into substreams of length $W = 2^{72}$. The state (and seed) of each stream is a vector of six 31-bit integers. This size of state is appropriate for having streams running in work-items on GPU cards, for example, while providing a sufficient period length for most applications. The initial state of the first stream for the MRG31k3p is by default a vector of six integers ``12345''.

>>>>

Graphics Processing Units (GPUs) have the potential to make important contribution to parallel computing with \proglang{R}. GPUs can perform thousands of tests simultaneously, which makes them powerful for doing massively parallel computing, and they are relatively cheap compared to multicore CPU's. Although there have been a number of \proglang{R} packages developed which provide some GPU capability, they inevitably come with some limitations. Packages such as \pkg{gputools}, \pkg{gpumatrix}, \pkg{cudaBayesreg}, \pkg{rpud} (available on github), are now no longer maintained, the popular \pkg{tensorflow} \citep{tensorflow1} package uses GPU via \proglang{Python}, which makes it difficult to include as a dependency for new \proglang{R} packages. All of these mentioned packages are restricted to \proglang{R} users with NVIDIA GPUs. \pkg{gpuR} \citep{gpur1} is the only \proglang{R} package with a convenient and flexible interface between \proglang{R} and GPUs, and it is compatible with many GPU devices. By utilizing the \pkg{ViennaCL} \citep*{rupp2016viennacl} library, it provides a bridge between \proglang{R} and non-proprietary GPUs through the \proglang{OpenCL} (Open Computing Language) backend, which when combined with \pkg{Rcpp} \citep{rcpp1} gives a building block for other \proglang{R} packages.

Random number generation is critical in simulation-based statistical inference, machine learning and many other scientific fields. While most random number generators are sequential, the \proglang{R} packages \pkg{parallel}, \pkg{rlecuyer} \citep{sevcikova2015package} and \pkg{future} \citep*{future1.19.1} are able to generate random numbers in parallel on multicore CPUs. More specifically, \pkg{parallel} writes an interface for the \pkg{RngStreams}, a \proglang{\texttt{C++}} library by \cite{l2002object} which is based on a combined multiple-recursive generator (MRG). \pkg{future} also uses the combined MRG algorithm for generating random numbers. For up-to-date review papers on the generation of random numbers on parallel devices, and GPUs in particular, see:  \cite{rLEC15a,rLEC17p,rLEC21a}.

The \pkg{clrng} package described here is currently the only \proglang{R} package that is able to generate random numbers on GPUs. Accomplishing this is complicated to do because each process must produce an independent sequence of random numbers, and in order to ensure reproducibility it should be possible to save and restore the current state of each stream of numbers at any point.  Here we introduce the \proglang{R} package \pkg{clrng} that leverages the \pkg{gpuR} package, and can generate random numbers on GPU by using the \pkg{clRNG} \citep{l2015clrng} library, and can thus accelerate several types of statistical simulation and modelling.

@@

\section{Uniform random number generation} \label{}
Uniform random number generators (RNGs) are the foundation for simulating random numbers from all types of probability distributions. \cite{l2012random} summarized the usual two steps to generate a random variable in computational statistics: (1) generating independent and identically distributed (i.i.d.) uniform random variables on the interval $(0, 1)$, (2) applying transformations to these i.i.d. $U(0, 1)$ random variables to get samples from the desired distribution. \cite{l2012random} and \cite{robert2004random} present many general transformation methods for generating non-uniform random variables, for example, the most frequently used inverse transform method, the Box-Muller algorithm \citep{box1958note} for Gaussian random variable generation, and so on, which are all built on uniform random variables. %In the rest paper, ``RNG'' refers to uniform pseudo random number generator.

\pkg{clRNG} is an \proglang{OpenCL} library for uniform random number generation, it provides four different RNGs: the MRG31k3p, MRG32k3a, LFSR113, and Philox-4×32-10 generators. These four RNGs use different types of constructions. \pkg{clrng} package uses the MRG31k3p RNG, making it able to generate random numbers on GPUs. We choose MRG31k3p as the base generator for the following reasons:  the original \pkg{RNGStreams} package \citep{l2002object} was built with MRG32k3a, which was designed to be implemented in double precision, and not with 32-bit integers. The MRG31k3p generator was designed later, specifically for 32-bit integer arithmetic, so it runs faster on the 32-bit GPUs.  It is also faster than Philox-4×32-10 \citep{rLEC00b}. MRG31k3p was also statistically tested extensively and successfully, \cite[see][]{rLEC07b}.

@@

\subsection{Creating streams}\label{createstreams}

THE PROBLEM:  each process needs to be independent, there's a risk of overlap.  solution: streams.  mrg31kp3 a stream is a vector of three positive integers.  initial state is first stream, compute new streams sequentially.

The popular way of obtaining multiple streams is to take an RNG with a long period and cut the RNG's sequence into very long pieces, these pieces are adjacent and of equal length $Z$, so each piece can be used as a separate stream. Creating a new stream amounts to computing its starting point (seed). Each of these streams can also be partitioned into substreams with equally-spaced starting points \citep{l2002object, rLEC15a}, which are occasionally useful.

In \proglang{clRNG} library, for the MRG31k3p RNG, the entire period length of approximately $2^{185}$ is divided into approximately $2^{51}$ non-overlapping streams of length $Z = 2^{134}$. Each stream can be further partitioned into substreams of length $W = 2^{72}$, although this is not currently implement in \pkg{clrng}. The state (and seed) of each stream is a vector of six 31-bit integers. This size of state is appropriate for having streams running in work-items on GPU cards, for example, while providing a sufficient period length for most applications. The initial state of the first stream for the MRG31k3p is by default $(12345, 12345, 12345)$.

@@
